Measuring the double-drift illusion with hand tracking

### 2021-03-01 Manuscript.Rmd
combined graphs, improved text of the manuscript

---
title: "Measuring the double-drift illusion with hand tracking"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

In experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 s, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 s to do so. Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 cps. Since each moving gabor was shown for 12 s, and it moved from one end to another in 2 s, it completed 5 full and 2 half passes of the path.

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds. The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking does not prevent the illusion from occurring. We will use heading angle as a measure of the strength of the illusion.

```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2. Heading and Trajectories in Experiment 1.** **Top:** Average heading across the middle 1 _s_ of each pass. Distance from origin is time, angle is angle. In darker blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading. **Bottom:** All individual trajectores. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity."}
plotBoundedTracking()
```

@@

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? We compare mean illusion strength to that reported previously [@Cavanagh2019] in Figure 3. As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model (in gray). That is; tracking the illusion while observing it does not change the strength of the illusion.

```{r fig3, echo=FALSE, message=FALSE, fig.width=5, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants. The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here."}
plotStrengthComparison()
```

@@

The durations of each pass in the online tracking experiment was 2 seconds but the manual traces mostly showed a linear trajectory without saturation or reset of the types shown in Figure 4 on the left below. The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks. To address this, we switched to a delayed, offline recording of the perceived path.

```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** **Left:** At some point, the perceived position will stop moving further away from its actual path. For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two. **Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate. **Bottom right:** if they occur instead after some period of time, they should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

@@

In Figure 4, we show the distributions of direction halfway to the reset points as well as trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes. This exemplary participant's re-traced paths show clear 'jump' resets, all of which are detected by our simple algorithm. Other participants are not that clear cut.

```{r fig5, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5. Illusion Strength and Example Trajectoriesin Experiment 2.** _Left_: For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show the distribution of angles for points halfway between the start point and detected resets; our measure of illusion strength. Dots below the distribution are individual participants' average illusion strengths. Gray lines indicate no illusion. _Right_: One exemplary participant's trajectories with very clear 'jump' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear."}
plotExampleData()
```

On average, the algorithm detects resets in ~75% of trials. For three participants, only about half the trials show resets (42%, 44% and 58%). For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with the average at ~17. Since participants experiencing a "hit-the-wall reset" (as opposed to a "jump" reset) might not always display a direction change in their paths, 50% of trials with resets seems reasonable. Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion. Therefore, we keep all participants and resets.

First, we consider the line angle responses as a measure of illusion strength. The strength of the illusion should depend on both the internal and external speed of the double drift stimulus. As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles. To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors. We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction. That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.cap="\\label{fig:fig5}**Figure 5. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as our main measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will therefore only use the angle of samples half-way to detected reset points as measure of illusion strength.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of the standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6. Reset Points and Single Limits** **Left:** True path of the double-drift stimulus as gray arrow. The average reset points for each participant and stimulus condition displayed spatially, with the internal motion normalized to go rightward. Y-coordinates have been converted to time ( _s_ ), in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

@@

Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope. If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.8) and a spatial limit of ~1.7 _cm_ (AIC=-208.6). This is close to the median values of the respective coordinates. The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs). This again indicates that a spatial limit explains more of the data.

@@

Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).

```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
plotSequentialModel()
```

The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. An as of yet unpublished study [@Liu2019] obtained data on the PSE of a double-drift stimulus at the end of trajectories of various lengths traveled at different speeds. These PSEs were predicted very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_). Interestingly, this temporal constraint fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then we should see a similar pattern in the averaged trajectories in our re-tracing experiment. However, these averaged trajectories reach a plateau within ~1 _s_ after which resets and illusory position shift seem to balance each other. This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. Our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias. But primarily, the PSE-based approach relies on all participants to exhibit full jump-resets, whereas the approach presented here is also valid for weaker resets. Nevertheless, we can't reolve this contradiction here and leave that for future work.

@@

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead usually triggered by the stimulus and not by attention. This should however be tested with eye-tracking data, so that this will remain a speculation here.

>>>>

---
title: "Measuring the double-drift illusion with hand tracking"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

In experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 s, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 s to do so. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 cps. Since each moving gabor was shown for 12 s, and it moved from one end to another in 2 s, it completed 5 full and 2 half passes of the path.

@@

In Figure 2 (top right), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds. The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking does not prevent the illusion from occurring. We will use heading angle as a measure of the strength of the illusion.

```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2. Online Tracking Data** **Left:** Individual trajectories. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity. **Top right:** Average heading in the middle 1 _s_ of each pass. Distance from origin is time, angle is angle. Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading. **Bottom right: The strength of the illusion in degrees angle deviation from the physical path over unity slope (internal / external velocity). In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants. Gray dashed line: unity. Black line: fit from Cavanagh & Tse (2019), 81% slope. Purple: fit to the data, 72% slope. **"}
plotBoundedTracking()
```

@@

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Cavanagh2019]. As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fit to previous data (in black). That is; tracking the illusion while observing it does not change the strength of the illusion.

@@

The durations of each pass in the online tracking experiment was 2 seconds but the manual traces (Figure 2, left) mostly showed a linear trajectory without saturation or reset of the types shown in Figure 3 on the left below. The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks. To address this, we switched to a delayed, offline recording of the perceived path.

```{r fig3, echo=F, fig.height=4.55, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3: Resets.** **Left:** At some point, the perceived position will stop moving further away from its actual path. For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two. **Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate. **Bottom right:** if they occur instead after some period of time, they should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

@@

In Figure 4 (left), we show the trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes. This exemplary participant's re-traced paths show clear 'jump' resets, all of which are detected by our simple algorithm. Other participants are not that clear cut.

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4. Data in Experiment 2.** **Left:*** One exemplary participant's trajectories with very clear 'jump' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear. **Middle:** Average participant reset points with standard deviations. (Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 3; the acrtangent of the slope (internal velocity over external velocity). Gray dahsed line: unity, black line: data/model as in Cavanagh & Tse (2019) at 81% of unity, purple line: best fit here at 42% of unity."}
#plotExampleData()
plotData()
```

On average, the algorithm detects resets in ~75% of trials. For three participants, only about half the trials show resets (42%, 44% and 58%). For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with the average at ~17. Since participants experiencing a "hit-the-wall reset" (as opposed to a "jump" reset) might not always display a direction change in their paths, 50% of trials with resets seems reasonable. Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion. Therefore, we keep all participants and resets (reset points in Figure 4, middle).

We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019]. A line through the origin with slope 0.42 predicts illusion strength fairly well in this data, which is weaker than earlier (0.81) or with online tracking (0.72). We can also see that illusion strength measured with re-tracing is more variable. While the illusion is weaker, it is nevertheless there and depends on the relationship between internal and external speed, as expected.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 4 (middle), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of the standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="\\label{fig:fig5}**Figure 5. Reset Models** **Left:** X-coordinates predicted from a single temporal limit. **Middle:** Y-coordinates predicted from a single spatial limit. **Right:** X- and Y-coordinates predicted from a temporal and a spatial limit."}
#plotResetPoints()
plotModels()
```

@@

Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope. If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.8) and a spatial limit of ~1.7 _cm_ (AIC=-208.6), as illustrated in Figure 6 (left and middle). This is close to the mean values of the respective coordinates (Figure 5, middle). The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs). This again indicates that a spatial limit explains more of the data.

## Combined Limits

@@

Figure 5 (right) shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).

The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the AIC of the model (-209.8) is slightly lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. An as of yet unpublished study [@Liu2019] obtained data on the PSE of a double-drift stimulus at the end of trajectories of various lengths traveled at different speeds. These PSEs were predicted very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_). Interestingly, this temporal constraint fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then we should see a similar pattern in the averaged trajectories in our re-tracing experiment. However, these averaged trajectories reach a plateau within ~1 _s_ after which resets and illusory position shift seem to balance each other. This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. Our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias. But primarily, the PSE-based approach relies on all participants to exhibit full jump-resets, whereas the approach presented here is also valid for weaker resets. Nevertheless, we can't resolve this contradiction here and leave that for future work.

@@

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead usually triggered by the stimulus and not by attention. This should however be tested with eye-tracking, so that this will remain a speculation here.
